What to simulate? Inferring the right direction for mental rotation

### 2014-02-07 man/cogsci-2014/cogsci-2014.tex
Tightening, etc.

% space below "Figure 1: ...", but only for inline figures
\addtolength{\textfloatsep}{-0.5cm}

\addtolength{\abovecaptionskip}{-0.25cm}
\addtolength{\belowcaptionskip}{-0.2cm}

@@

\begin{abstract}
  When people use mental imagery, how do they decide \textit{which}
  images to generate? To answer this question, we explored how mental
  simulation should be used in the classic psychological task of
  determining if two images depict the same object in different
  orientations \cite{Shepard1971}. Through a rational analysis of
  mental rotation, we formalized four models and compared them to
  human performance. \TODO{None of the models were a particularly
    satisfying explanation of human behavior, leading us to suggest
    that the question of ``what to simulate?'' is more difficult than
    has previously been assumed. While only exploratory, this work
    provides the foundation necessary to begin answering this
    question.}

@@

One of the most astonishing cognitive feats is our ability to
envision, manipulate, and plan with objects -- all without actually
perceiving them. This capacity for ``mental simulation'' has been
widely studied, including an intense debate about the underlying
representation of mental images \cite<e.g.,>{Kosslyn:2009tj,
  Pylyshyn:2002vk}. But this debate hasn't addressed one of the most
fundamental questions about mental simulation: how people decide what
they should simulate.

Mental rotation provides a simple example of the decision problem
posed by simulation.  In the classic experiment by
\citeA{Shepard1971}, participants viewed images of three-dimensional
objects and had to determine whether the images depicted the same
object (which differed by a rotation) or two separate objects (which
differed by a reflection and a rotation). They found that people's
response times had a strong linear correlation with the minimum angle
of rotation, a result which led to the conclusion that people solve
this task by ``mentally rotating'' the objects until they are
congruent.  However, this explanation leaves several questions
unanswered. How do people know the axis around which to rotate the
objects? If the axis is known, how do people know which direction to
rotate the objects?  And finally, how do people know how long to
rotate?

@@

The original ``congruency'' hypothesis \cite{Shepard1971} actually is
a rational solution to this problem, in the sense that the smallest
amount of computation coincides with rotating through the minimum
angle. However, it violates the constraint that we do not know the
points of correspondence between the images, which is exactly what
necessitates the use of imagery. Noting that a rational solution need
not necessarily maintain a single trajectory of rotation, we explore
an alternative rational model, which--rather than computing the angle
of rotation ahead of time--engages in an \textit{active sampling}
strategy.

@@

Additionally, we assume that the observed shapes may only be
transformed by a small amount at a time, and each transformation is
costly in terms of computational resources. Specifically, if the
current mental image is $\Xt$, then:
\begin{equation}
  \mathbf{X}_{t+1} = \left\{ \begin{array}{ll}
      f(\Xt, 0, \epsilon) &\mbox{ continue rotating by $\epsilon$,} \\
      f(\Xt, 1, 0) &\mbox{ flip,} \\
      f(\Xa, 0, 0) &\mbox{ reset, or} \\
      f(\Xa, 1, 0) &\mbox{ reset and flip,} \\
    \end{array} \right.
  \label{eq:actions}
\end{equation}
where $\epsilon\sim \left|\mathcal{N}(0, \sigma_\epsilon^2)\right|$
and $\sigma_\epsilon^2$ is the variance of the step size in radians.

@@

In order to approximate Equation~\ref{eq:lh-ratio} using samples of
the similarity function, we must decide \textit{which} places to
sample and \textit{when} stop sampling. The models below differ in how
they make these decisions (see Figure~\ref{fig:model-traces} for
examples of their behavior).

@@

To determine the correct rotation, we solve for the rotation matrix by
computing $(\Xa \F_h^T)_\mathrm{left}^{-1}\cdot{}\Xb$, where
$(\Xa\F_h^T)_\mathrm{left}^{-1}$ is the left inverse of
$\Xa\F_h^T$. We then check each $h$ to see if the computation produces
a valid rotation matrix; the $h$ that does is the correct
hypothesis. This approach gives us the true value of $\theta$, so
Equation~\ref{eq:lh-ratio} becomes a generalized likelihood ratio
test, where $\theta$ is set to the MLE value, rather than being
marginalized:
\begin{equation}
  \ell = \frac{\max_\theta p(\Xb\ \vert\ \Xa, \theta, \hi)p(\hi)}{\max_\theta p(\Xb\ \vert\ \Xa, \theta, \hf)p(\hf)}.
  \label{eq:mle-lh-ratio}
\end{equation}

@@

\begin{figure*}[t]
  \begin{center}
    \includegraphics[width=\textwidth]{../../figures/D/response_time_scatters.pdf}
    \caption{\textbf{Model vs. human RTs.} Each subplot shows the
      z-scored model response times ($x$-axis) vs. the z-scored human
      response times ($y$-axis). Pearson correlations are shown
      beneath each subplot. The dotted lines are $x=y$.}
    \label{fig:human-model-scatters}
  \end{center}
\end{figure*}

In the current formulation of the problem, choosing the threshold is
straightforward because we know both the exact geometry of the shapes
and that a linear transformation exists which will perfectly align
them. However, this choice is not always clear \textit{a priori}, as
the global optimum depends on many factors (e.g., shape complexity,
dimensionality, perceptual uncertainty, and whether the shapes are
identical).  One way to deal with the problem of choosing a threshold
would simply be to perform an exhaustive search or use a global
optimization strategy. However, these strategies would not result in
the linear response time found by \citeA{Shepard1971}. A second
alternative, which could potentially demonstrate the linear response
time result, might instead just perform a Hill Climbing (\Hc{}) search
for a \textit{local} maximum. As with the \Oc{} and \Th{} models,
Equation~\ref{eq:lh-ratio} becomes a generalized likelihood ratio
test.

@@

Denoting $S_h=S(\Xb, f(\Xa, \theta, h))$, we first place a
\textit{Gaussian Process} \cite{Rasmussen:2006vz}, or GP, prior on the
log of $S_h$ in order to enforce positivity after it is exponentiated,
i.e. $\mathbb{E}[Z_h] \approx \int
\exp(\mu_h(\theta))p(\theta)\dif\theta$, where $\mu_h:=\mu(\log S_h)$
is the mean of the log-GP \cite{Osborne:2012tm}.  However, this
integral is still intractable. To approximate it, we fit a second GP
over points sampled from the log-GP, which we denote as
$\bar{S}_h:=\exp(\mu_h)$.\footnote{Note that the approach used here
  differs slightly from the one presented in \citeA{Osborne:2012tm}
  and is based on personal communication with one of the authors
  \cite{Duvenaud:2013td}.} Then, from \citeA{Duvenaud:2013td}:
\begin{align}
  \mathbb{E}[Z_h] &\approx \int \bar{\mu}_h(\theta)p(\theta)\dif\theta,\\
  \mathbb{V}(Z_h) &\approx \iint \mathrm{Cov}_h(\theta,
  \theta^\prime)\bar{\mu}_h(\theta)\bar{\mu}_h(\theta^\prime)p(\theta)p(\theta^\prime)\dif\theta\dif\theta^\prime,
\end{align}
where $\bar{\mu}_h:=\mu(\bar{S}_h)$ is the mean of the second GP, and
$\mathrm{Cov}_h:=\mathrm{Cov}(\log S_h)$ is the covariance of the
log-GP.

@@

While there was no limit on response time, participants were urged to
respond as quickly as possible while still being
accurate. Specifically, we asked them to aim for at least 85\%
accuracy in the experimental blocks, and provided a counter to keep
them informed of their score.

@@

Out of the 247 participants, 200 (81\%) were included in our
analyses. Of the rest, we excluded 10 (4\%) because of an experimental
error, 6 (2.4\%) because they had already completed a related
experiment, and 31 (12.6\%) because they failed a comprehension
check. Comprehension was defined as correctly answering at least 85\%
of stimuli with a minimum rotation of $0^\circ$ or $20^\circ$.
Additionally, we excluded 82 trials for which the response time (RT)
was either less than 100 milliseconds or greater than 20 seconds.

For each model, we ran 50 samples for each of the 800 experimental
stimuli. The step size parameter ($\sigma_\epsilon$) was fit to human
RTs for each of the models, resulting in $\sigma_\epsilon=0.6$ for the
\Th{} and \Bq{} models and $\sigma_\epsilon=0.1$ for the \Oc{} and
\Hc{} models. We also ran the models under two different priors,
$p(h=0)=0.5$ (the ``equal'' prior) and $p(h=0)=0.55$ (the ``unequal''
prior). This only had a major effect on the stopping criteria for the
\Bq{} model.\footnote{The prior also had a small effect on the
  accuracy of the \Hc{} model, because sometimes the local maxima were
  small enough that Equation~\ref{eq:mle-lh-ratio} became dominated by
  the prior.}

@@

The average RT across all correctly-judged stimuli was \ExpTime{}
milliseconds; the full histogram of response times can be seen in
Figure~\ref{fig:histograms}. The minimum angle of rotation was
significantly rank-order (Spearman) correlated with average
per-stimulus RTs, both for ``flipped'' (\ExpThetaTimeCorrFlipped{})
and ``same'' pairs (\ExpThetaTimeCorrSame{}). While this replicates
the general result of previous experiments
\cite<e.g.,>{Shepard1971,Cooper:1975wp}, our results are not as
perfectly linear.

@@

There was an overall monotonic relationship between the minimum angle
of rotation and the number of actions taken by the \Th{} model
(Figure~\ref{fig:response-time-accuracy}). However, the relationship
between angle and RT for individual shapes was usually not monotonic
(e.g., Figure~\ref{fig:response-time-stimulus}).  The \Th{} model was
able to explain a moderate amount of the variance in human RTs, with a
correlation of \ThTimeCorr{}
(Figure~\ref{fig:human-model-scatters}). Like the \Oc{} model,
however, the overall distribution of its response times did not match
that of people (Figure~\ref{fig:histograms}). Additionally, the \Th{}
model had 100\% accuracy, and thus did not exhibit a relationship
between minimum angle and accuracy.

As noted above, we fit $\sigma_\epsilon=0.6$ for the \Th{} model. This
has the interesting effect of actually causing the \Th{} to
\textit{over}rotate (see, for example, the ``flipped'' pair trace in
Figure~\ref{fig:model-traces}). This is because the step size is large
enough that it sometimes misses the global maximum, and must do
another full rotation to find it. However, when the step size is small
enough to always find the maximum, the fit to human data is slightly
worse.

@@

The \Oc{} model was the best predictor of human RTs, but is an
unsatisfying explanation of human behavior because it relies on
\textit{a priori} knowledge that people are unlikely to have about
which points on the two shapes correspond. Similarly, while the \Th{}
was next best in terms of correlation, it makes the unsatisfying
assumption that the global maximum is known. Finally, neither of these
models can explain the difference in people's behavior on ``same'' and
``flipped'' stimuli.  The \Bq{} model was nearly as good as the \Th{}
model; in contrast to the other models, the \Bq{} model makes no
assumptions about people's \textit{a prior} knowledge. Furthermore,
the \Bq{} model's adaptive stopping rule is sensitive to the prior,
and thus provides a possible explanation for why people are slower to
respond on ``flipped'' stimulus pairs.

Additionally, a closer look at performance on individual shapes
reveals several interesting points
(Figure~\ref{fig:response-time-stimulus} shows curves for one
particular shape; these curves are fairly typical relative to other
shapes). First, the curves for the human RTs on specific shapes tend
to either be linear, or linear first, then flat. Second, while the
\Th{} model appears to capture the concavity of human RTs overall, it
is actually fairly inconsistent within shapes. Third, the linearity of
the \Bq{} model's RT curves \textit{is} consistent within shapes,
though it does not exhibit the same ``plateau'' effect that sometimes
appears in the human curves.

Thus, we suggest that the \Bq{} model offers the most promising
explanation of people's behavior on the mental rotation task to
date. It is the only model out of those surveyed which did not make
assumptions about people's \textit{a priori} knowledge, which was a
reasonable predictor of people's behavior, and which provided an
explanation for the longer response times on ``flipped'' stimuli.
Clearly, though, the \Bq{} model is not a perfect account, and there
are several immediately obvious ways in which it could be
improved. For example, while we used holistic rotations in this paper,
there is evidence that people compare individual features of shapes
rather than the entire shape at once
\cite{Just1976,Yuille:1982tx}. Second, there are other possible active
sampling approaches that could combine the best features from the
\Th{} and \Bq{} models, such as maintaining a distribution over the
location and value of the global maximum, rather than over the
integral.

>>>>

% space below "Figure 1: ...", but only for inline figures
%\addtolength{\textfloatsep}{-0.5cm}

\addtolength{\abovecaptionskip}{-0.25cm}
\addtolength{\belowcaptionskip}{-0.3cm}

@@

\begin{abstract}
  When people use mental imagery, how do they decide \textit{which}
  images to generate? To answer this question, we explored how mental
  simulation should be used in the classic psychological task of
  determining if two images depict the same object in different
  orientations \cite{Shepard1971}. Through a rational analysis of
  mental rotation, we formalized four models and compared them to
  human performance. We found that three of the models, which were
  based on previous hypotheses in the literature, were unable to
  account for several aspects of human behavior. The fourth, which is
  a model based on the idea of \textit{active sampling}
  \cite<e.g.,>{Gureckis:2012gu}, provides plausible explanations where
  the other models do not. Based on these results, we suggest that the
  question of ``what to simulate?'' is more difficult than has
  previously been assumed, and that an active learning approach holds
  promise for uncovering the answer.

@@

One of the most astonishing cognitive feats is our ability to
envision, manipulate, and plan with objects---all without actually
perceiving them. This \textit{mental simulation} has been widely
studied, including an intense debate about the underlying
representation of mental images \cite<e.g.,>{Kosslyn:2009tj,
  Pylyshyn:2002vk}. But this debate hasn't addressed one of the most
fundamental questions about mental simulation: how people decide what
they should simulate.

Mental rotation provides a simple example of the decision problem
posed by simulation.  In the classic experiment by
\citeA{Shepard1971}, participants viewed images of three-dimensional
objects and had to determine whether the images depicted the same
object (which differed by a rotation) or two separate objects (which
differed by a reflection and a rotation). They found that people's
response times (RTs) had a strong linear correlation with the minimum
angle of rotation, a result which led to the conclusion that people
solve this task by ``mentally rotating'' the objects until they are
congruent.  However, this explanation leaves several questions
unanswered. How do people know the axis around which to rotate the
objects? If the axis is known, how do people know which direction to
rotate the objects?  And finally, how do people know how long to
rotate?

@@

The original ``congruency'' hypothesis \cite{Shepard1971} actually is
a rational solution to this problem, in the sense that the smallest
amount of computation coincides with rotating through the minimum
angle. However, it violates the constraint that we do not know the
points of correspondence between the images, which is exactly what
necessitates the use of imagery. Noting that a rational solution need
not necessarily maintain a single trajectory of rotation, we explore
an alternative model, which---rather than explicitly computing the
angle of rotation---engages in an \textit{active sampling} strategy.

@@

Additionally, we assume that the observed shapes may only be
transformed by a small amount at a time, and each transformation is
costly in terms of computational resources. Specifically, if the
current mental image is $\Xt$, then:
\begin{equation}
  \mathbf{X}_{t+1} = \left\{ \begin{array}{ll}
      f(\Xt, 0, \epsilon) &\mbox{ rotate by $\epsilon$ radians,} \\
      f(\Xt, 1, 0) &\mbox{ flip,} \\
      f(\Xa, 0, 0) &\mbox{ reset, or} \\
      f(\Xa, 1, 0) &\mbox{ reset and flip,} \\
    \end{array} \right.
  \label{eq:actions}
\end{equation}
where $\epsilon\sim \left|\mathcal{N}(0, \sigma_\epsilon^2)\right|$
and $\sigma_\epsilon^2$ is the variance of the step size.

@@

In order to approximate Equation~\ref{eq:lh-ratio} using samples of
the similarity function, we must decide \textit{which} places to
sample and \textit{when} stop sampling. The models below differ in how
they make these decisions; see Figure~\ref{fig:model-traces} for an
example.

@@

To determine the correct rotation, we solve for the rotation matrix by
computing $(\Xa \F_h^T)_\mathrm{left}^{-1}\cdot{}\Xb$, where
$(\Xa\F_h^T)_\mathrm{left}^{-1}$ is the left inverse of
$\Xa\F_h^T$. We then check each $h$ to see if the computation produces
a valid rotation matrix; the $h$ that does is the correct
hypothesis. This gives us the true value of $\theta$, so
Equation~\ref{eq:lh-ratio} becomes a generalized likelihood ratio
test, where $\theta$ is set to the MLE value, rather than being
marginalized:
\begin{equation}
  \ell = \frac{\max_\theta p(\Xb\ \vert\ \Xa, \theta, \hi)p(\hi)}{\max_\theta p(\Xb\ \vert\ \Xa, \theta, \hf)p(\hf)}.
  \label{eq:mle-lh-ratio}
\end{equation}

@@

\begin{figure*}[t]
  \begin{center}
    \includegraphics[width=\textwidth]{../../figures/D/response_time_scatters.pdf}
    \caption{\textbf{Model vs. human RTs.} Each subplot shows the
      z-scored model RTs ($x$-axis) vs. the z-scored human RTs
      ($y$-axis). Pearson correlations are shown beneath each
      subplot. The dotted lines are $x=y$.}
    \label{fig:human-model-scatters}
  \end{center}
\end{figure*}

In the current formulation of the problem, choosing the threshold is
straightforward because we know both the exact geometry of the shapes
and that a linear transformation exists which will perfectly align
them. However, this choice is not always clear \textit{a priori}, as
the global optimum depends on many factors (e.g., shape complexity,
dimensionality, perceptual uncertainty, and whether the shapes are
identical).  One way to deal with the problem of choosing a threshold
would simply be to perform an exhaustive search or use a global
optimization strategy. However, these strategies would not result in
the linear RT found by \citeA{Shepard1971}. A second alternative might
instead just perform a Hill Climbing (\Hc{}) search for a
\textit{local} maximum. As with the \Oc{} and \Th{} models,
Equation~\ref{eq:lh-ratio} becomes a generalized likelihood ratio
test.

@@

Denoting $S_h=S(\Xb, f(\Xa, \theta, h))$, we first place a
\textit{Gaussian Process} \cite{Rasmussen:2006vz}, or GP, prior on the
log of $S_h$ in order to enforce positivity after it is exponentiated,
i.e. $\mathbb{E}[Z_h] \approx \int
\exp(\mu_h(\theta))p(\theta)\dif\theta$, where $\mu_h:=\mu(\log S_h)$
is the mean of the log-GP \cite{Osborne:2012tm}.  However, this
integral is still intractable. To approximate it, we fit a second GP
over points sampled from the log-GP, which we denote as
$\bar{S}_h:=\exp(\mu_h)$.\footnote{Note that the approach used here
  differs slightly from the one presented in \citeA{Osborne:2012tm}
  and is based on personal communication with one of the authors
  \cite{Duvenaud:2013td}.} Then, from \citeA{Duvenaud:2013td}, we have
$\mathbb{E}[Z_h] \approx \int \bar{\mu}_h(\theta)p(\theta)\dif\theta$
and $\mathbb{V}(Z_h) \approx \iint \mathrm{Cov}_h(\theta,
\theta^\prime)\bar{\mu}_h(\theta)\bar{\mu}_h(\theta^\prime)p(\theta)p(\theta^\prime)\dif\theta\dif\theta^\prime$,
where $\bar{\mu}_h:=\mu(\bar{S}_h)$ is the mean of the second GP, and
$\mathrm{Cov}_h:=\mathrm{Cov}(\log S_h)$ is the covariance of the
log-GP.

@@

While there was no limit on RT, participants were urged to respond as
quickly as possible while still being accurate. Specifically, we asked
them to aim for at least 85\% accuracy in the experimental blocks, and
provided a counter to keep them informed of their score.

@@

Out of the 247 participants, 200 (81\%) were included in our
analyses. Of the other 47, we excluded 10 (4\%) because of an
experimental error, 6 (2.4\%) because they had already completed a
related experiment, and 31 (12.6\%) because they failed a
comprehension check. Comprehension was defined as correctly answering
at least 85\% of stimuli with a minimum rotation of $0^\circ$ or
$20^\circ$.  Additionally, we excluded 82 trials for which the RT was
either less than 100 milliseconds or greater than 20 seconds.

For each model, we ran 50 samples for each of the 800 experimental
stimuli. The step size parameter ($\sigma_\epsilon$) was fit to human
RTs for each of the models, resulting in $\sigma_\epsilon=0.6$ for the
\Th{} and \Bq{} models and $\sigma_\epsilon=0.1$ for the \Oc{} and
\Hc{} models. We also ran the models under two different priors,
$p(h=0)=0.5$ (the ``equal'' prior) and $p(h=0)=0.55$ (the ``unequal''
prior). This only had a major effect on the stopping criteria for the
\Bq{} model.

@@

The average RT across all correctly-judged stimuli was \ExpTime{}
milliseconds; the full histogram of RTs can be seen in
Figure~\ref{fig:histograms}. The minimum angle of rotation was
significantly rank-order (Spearman) correlated with average
per-stimulus RTs, both for ``flipped'' (\ExpThetaTimeCorrFlipped{})
and ``same'' pairs (\ExpThetaTimeCorrSame{}). While this replicates
the general result of previous experiments
\cite<e.g.,>{Shepard1971,Cooper:1975wp}, our results are not as
perfectly linear.

@@

There was an overall monotonic relationship between the minimum angle
of rotation and the number of actions taken by the \Th{} model
(Figure~\ref{fig:response-time-accuracy}). However, the relationship
between angle and RT for individual shapes was usually not monotonic
(e.g., Figure~\ref{fig:response-time-stimulus}).  The \Th{} model was
able to explain a moderate amount of the variance in human RTs, with a
correlation of \ThTimeCorr{}
(Figure~\ref{fig:human-model-scatters}). Like the \Oc{} model,
however, the overall distribution of its RTs did not match that of
people (Figure~\ref{fig:histograms}). Additionally, the \Th{} model
had 100\% accuracy, and thus did not exhibit a relationship between
minimum angle and accuracy.

As noted above, we fit $\sigma_\epsilon=0.6$ for the \Th{} model. This
has the interesting effect of actually causing the \Th{} to
\textit{over}rotate (see, for example, the ``flipped'' pair trace in
Figure~\ref{fig:model-traces}). This is because the step size is large
enough that it sometimes misses the global maximum, and must do
another full rotation to find it. However, when the step size is
smaller, the fit to human data is slightly worse.

@@

The \Oc{} and \Th{} models were---in terms of correlations---the best
predictors of human RTs. However, both are somewhat unsatisfying
explanations because they rely on \textit{a priori} knowledge that
people are unlikely to have. Moreover, they offer no explanation of
several aspects of human behavior. First, their overall RT
distributions look nothing like people's
(Figure~\ref{fig:histograms}).  Second, they both are 100\% accurate,
and so cannot explain the systematic relationship between rotation and
accuracy found in people
(Figure~\ref{fig:response-time-accuracy}). Third, neither model can
explain the difference in people's behavior on ``same'' and
``flipped'' stimuli.

In contrast, the \Bq{} model was nearly as good as the \Th{} model,
yet it makes no assumptions about people's \textit{a priori}
knowledge. Furthermore, the \Bq{} model matches people's behavior
better than the \Oc{} or \Th{} models in several ways. Its overall RT
histogram has the same general shape as people's
(Figure~\ref{fig:histograms}).  Moreover, a closer look shows that the
\Bq{} model maintains the monotonic relationship between angle and RT
even on individual stimuli, while the \Th{} model does not
(Figure~\ref{fig:response-time-stimulus}). Finally, the \Bq{} model's
adaptive stopping rule is sensitive to the prior, and thus provides a
possible explanation for why people are slower to respond on
``flipped'' stimulus pairs.

Thus, we suggest that the \Bq{} model offers the most promising
explanation of people's behavior on the mental rotation task to
date. While it is not a perfect account, there are several ways in
which it could be improved. For example, while we used holistic
rotations in this paper, there is evidence that people compare
individual features of shapes
\cite{Just1976,Yuille:1982tx}. Additionally, a different active
sampling approach could maintain a distribution over the location and
value of the global maximum, rather than over the integral. We intend
to explore these possibilities in future work, building upon the
foundation established in this paper and working towards a better
understanding of \textit{what} people should choose to simulate.
